---
title: "scSEQ analysis of the developing endothelium"
author: Nick Negretti
date: 11/17/20
output: rmarkdown::github_document
---

# Analysis of the lung endothelium

## Load libraries and helper functions
```{r, results="hide", message = FALSE}
setwd("~/postdoc/code/devo_scseq_github")
source("./helper_functions/globals.R")
source("./helper_functions/libraries.R")

opts_knit$set(root.dir = getwd())

source("./helper_functions/trajectory.R")
source("./helper_functions/cluster.R")
source("./helper_functions/colors.R")
source("./helper_functions/brackets.R")
source("./helper_functions/heatmaps.R")

plan("multiprocess", workers = N_WORKERS)
```
## Load data from merge and clean pipeline
```{r}
endo <- readRDS("./data/endo_full_noGm42418_sct_p7b_integrated_retransform.rds")
```


## UMAP and clustering
```{r, fig.width = 12}
plan("sequential")

# Make the tiempoints an ordered factor, so they are displayed consistantly
endo$timepoint <- ordered(as.factor(endo$timepoint), unique(endo$timepoint))

endo <- cluster_pca_umap(endo, k_param = 20, dims_umap = 1:20, dims_neighbors = 1:20, cluster_res = 0.1) # Note, this breaks if future is set to plan: multiprocess


p_cluster <- DimPlot(endo) + umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_category_20) +
        theme(aspect.ratio=1)
p_time <- DimPlot(endo, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_scanpy_default)+
        theme(aspect.ratio=1)

p_cluster + p_time + plot_annotation("Endothelium")

```


## Identify marker genes in each cluster
```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/endo_markers_clusters.rds"
if (!file.exists(filename)) {
  endo_markers <- parallelFindAllMarkers(endo)
  saveRDS(endo_markers, filename)
} else {
  endo_markers <- readRDS(filename)
}
```

### Cluster 0
```{r}
endo_markers[[0 + 1]][n_print,]
```

### Cluster 1
```{r}
endo_markers[[1 + 1]][n_print,]
```

### Cluster 2
```{r}
endo_markers[[2 + 1]][n_print,]
```

### Cluster 3
```{r}
endo_markers[[3 + 1]][n_print,]
```

### Cluster 4
```{r}
endo_markers[[4 + 1]][n_print,]
```

### Cluster 5
```{r}
endo_markers[[5 + 1]][n_print,]
```


## Use known marker genes to determine the identity of the clusters

Most of the markers were also identified: https://elifesciences.org/articles/53072
```{r, fig.width = 12}


marker_genes <- c("Gpihbp1", "Kit", # Microvascular
                  "Car4", "Kdr", # Car4
                  "Mki67", "Top2a", # Proliferating miEC
                  "Vwf", "Vcam1", #macrovascular
                  "Vegfc", "Prss23", #Venous macrovascular
                  "Cxcl12", "Pcsk5", #Artearial macrovascular
                  "Ephb4", # Vein
                  "Flt4", "Ccl21a" # Lymphatic
)


heatmap_df <- make_heatmap_df(endo, marker_genes, sort_clusters = FALSE)

heatmap_df <- heatmap_df %>% group_by(gene,cluster) %>% summarise(expression_mean=mean(expression))

# For brackets and text (see ./helper_functions/brackets.R)
n_clusters <- max(as.numeric(Idents(endo)))
text_vec <- list(c("Microvascular", 2),
                 c("Car4+ EC", 2),
                 c("Prolif.", 2),
                 c("maEC", 2),
                 c("Venous maEC", 2),
                 c("Arterial maEC", 2),
                 c("Venous", 1),
                 c("Lymphatic", 2)


)

endo_heatmap <- ggplot(heatmap_df, aes(x = gene, y = cluster, fill = expression_mean)) +
        geom_tile(color = "white", size = 0.1) +
        scale_fill_distiller(palette = "Blues", direction = 1, trans = "sqrt", name = "Expression") +
        coord_fixed(ratio = 1, xlim = NULL, ylim = c(1,n_clusters), expand = TRUE, clip = "off") +
        theme(plot.margin=unit(c(1.5,1,1,1),"cm")) +
        addBrackets(text_vec) +
        addText(text_vec, n_clusters) +
        theme(panel.background = element_rect(fill = "white", colour = "black", size = 0),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.title.x  = element_blank(),
              axis.text.x = element_text(size = 12, face = "italic"),
              axis.text.y = element_text(size = 12),
              axis.title.y = element_text(size = 14),
              legend.title = element_text(size = 14),
              legend.text = element_text(size = 12)
              ) +
        labs(y = "Cluster") +
        theme(axis.text.x = element_text(angle = 45, hjust=1))

endo_heatmap


saveTiff("./figures/endo/supp_cluster_markers.tiff",
         endo_heatmap,
         width = 12, height = 5)

```

## Save relabeled cell-types for future use

```{r}
endo_relabel <- endo
# Label the cell-types
levels(Idents(endo_relabel)) <- c("miEC", #0
                                  "Car4+ EC", #1
                                  "Proliferating miEC", #2
                                  "Venous maEC", #3
                                  "Arterial maEC", #4
                                  "Lymphatic" #5

)
Idents(endo_relabel) <- ordered(Idents(endo_relabel), c("Arterial maEC", "Venous maEC", "miEC", "Car4+ EC", "Proliferating miEC", "Lymphatic"))
endo_relabel$bulk_celltype <- "Endothelium"

filename <- "./data/endo_full_celltype_labels.rds"
if (!file.exists(filename)) {
  saveRDS(endo_relabel, filename)
} else {
  endo_relabel <- readRDS(filename)
}
```



## Plot relabaled cells
```{r, fig.width = 12}

renameFactorIdent <- function(obj, start, end){
  levels(obj)[levels(obj) == start] <- end
  return(obj)
}

Idents(endo_relabel) <- renameFactorIdent(Idents(endo_relabel), "Car4+ EC", "*Car4*+ EC")


p_rename_cluster <- DimPlot(endo_relabel) + umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_markdown(size=14),
              legend.title=element_text(size=14),
              plot.title = element_text(size = 14)
              )

p_time <- DimPlot(endo_relabel, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14),
              plot.title = element_text(size = 14))

p_time + p_rename_cluster + plot_annotation("Endothelium", theme = theme(plot.title = element_text(size = 14)))


saveTiff("./figures/endo/umap.tiff",
         p_rename_cluster + plot_annotation("Endothelium"),
         width = 6, height = 4.5)

saveTiff("./figures/endo/supp_umap.tiff",
         p_cluster + p_time + p_rename_cluster + plot_annotation("Endothelium", theme = theme(plot.title = element_text(size = 14))),
         width = 18, height = 4.5)

```

## Re-confirm the markers in the relabeled clusters

```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/endo_markers_clusters_relabeled.rds"
if (!file.exists(filename)) {
  endo_relabeled_markers <- parallelFindAllMarkers(endo_relabel)
  saveRDS(endo_relabeled_markers, filename)
} else {
  endo_relabeled_markers <- readRDS(filename)
}
```


### Arterial maEC (macrovascular)
```{r}
endo_relabeled_markers[[0 + 1]][n_print,]
```

### Venous maEC (macrovascular)
```{r}
endo_relabeled_markers[[0 + 2]][n_print,]
```
### miEC (microvascular)
```{r}
endo_relabeled_markers[[0 + 3]][n_print,]
```

### Car4+
```{r}
endo_relabeled_markers[[0 + 4]][n_print,]
```

### Lymphatic
```{r}
endo_relabeled_markers[[0 + 5]][n_print,]
```

### Proliferating miEC
```{r}
endo_relabeled_markers[[0 + 6]][n_print,]
```


```{r}
timepoint_factor <- factor(endo_relabel$timepoint)
timepoint_ordered <- ordered(timepoint_factor, levels = unique(endo_relabel$timepoint))

cell_ident_df <- data.frame(timepoint = timepoint_ordered,
                            cluster = Idents(endo_relabel))

prop_celltypes <- round(prop.table(table(cell_ident_df$timepoint, cell_ident_df$cluster), 1) * 100, 1)


endo_mountain <- ggplot(melt(prop_celltypes), aes(y=value, x=Var1)) +
        geom_area(aes(color = as.factor(Var2), group=as.factor(Var2), fill = as.factor(Var2))) +
        xlab("Day") +
        ylab("Percent of cells") +
        labs(fill = "Cluster", color = "Cluster", group = "Cluster") +
        scale_colour_manual(aesthetics = c("color", "fill"), values=color_category_20) +
        theme(legend.key=element_blank(),
              legend.text=element_markdown(size=14),
              legend.title=element_text(size=14),
              axis.text.x  = element_text(size=14),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=14),
              axis.title.y = element_text(size=14),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_rect(fill = "white")
        ) +
        scale_x_discrete(expand = c(.01, .01)) +
        scale_y_continuous(expand = c(.01,.01)) +
        coord_fixed(0.05)

endo_mountain


saveTiff("./figures/endo/labeled_mountain.tiff",
         endo_mountain,
         width = 7, height = 4)

```

## Car4 expression across the endothelium
```{r, fig.width = 6}
car4_endo_plot <- FeaturePlot(endo_relabel, "Car4") +
        featureplot_theme() +
        theme(aspect.ratio=1,
              plot.title = element_text(face = "italic", size = 14),
              legend.title = element_text(size = 14),
              legend.text = element_text(size = 12)
        ) +
        scale_color_viridis(name = "Expression", direction = -1)

car4_endo_plot

saveTiff("./figures/endo/endo_car4_plot.tiff",
         car4_endo_plot,
         width = 6, height = 4.5)

```

## Kdr expression across the endothelium
```{r, fig.width = 6}
kdr_endo_plot <- FeaturePlot(endo_relabel, "Kdr", pt.size = 0.5) +
        featureplot_theme() +
        theme(aspect.ratio=1,
              plot.title = element_text(face = "italic", size = 18),
              legend.title = element_text(size = 14),
              legend.text = element_text(size = 12)
        ) +
        scale_color_viridis(name = "Expression", direction = -1)

kdr_endo_plot

saveTiff("./figures/endo/endo_kdr_plot.tiff",
         kdr_endo_plot,
         width = 6, height = 4.5)

```

## Integrating velocity measures


```{r}
endo_relabel$car4_prob <- readRDS("./velocity/endo/data/endo_car4_prob.rds")$to_Car4
endo_relabel$car4_originators <- "no"
endo_relabel$car4_originators[endo_relabel$car4_prob > 0.2 & Idents(endo_relabel) != "Car4"] <- "yes"
```

```{r}
car4_originator_markers <- FindMarkers(endo_relabel, group.by = "car4_originators", "yes", "no", test.use = "MAST")
(car4_originator_markers %>% arrange(-avg_logFC))[n_print,]
```

```{r}
write.xlsx(car4_originator_markers, file = "./figures/endo/Table S2 - Commitment to Car4.xlsx", rowNames = TRUE)
```