---
title: "scSEQ analysis of the developing endothelium"
author: Nick Negretti
date: 11/17/20
output: rmarkdown::github_document
---

# Analysis of the lung endothelium

## Load libraries and helper functions
```{r, results="hide", message = FALSE}
setwd("~/postdoc/code/devo_scseq")

source("./helper_functions/libraries.R")

opts_knit$set(root.dir = getwd())

source("./helper_functions/trajectory.R")
source("./helper_functions/cluster.R")
source("./helper_functions/colors.R")
source("./helper_functions/brackets.R")
source("./helper_functions/heatmaps.R")
set.seed(42) # For reproducability

N_WORKERS <- 3
options(future.globals.maxSize=4*1024*1024^2)
plan("multiprocess", workers = N_WORKERS)
```
## Load data from merge and clean pipeline
```{r}
endo <- readRDS("./data/endo_full_noGm42418_sct_p7b_integrated_retransform.rds")
```


## UMAP and clustering
```{r, fig.width = 12}
plan("sequential")

# Make the tiempoints an ordered factor, so they are displayed consistantly
endo$timepoint <- ordered(as.factor(endo$timepoint), unique(endo$timepoint))

endo <- cluster_pca_umap(endo, k_param = 20, dims_umap = 1:20, dims_neighbors = 1:20, cluster_res = 0.1) # Note, this breaks if future is set to plan: multiprocess


p_cluster <- DimPlot(endo) + umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_category_20) +
        theme(aspect.ratio=1)
p_time <- DimPlot(endo, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_scanpy_default)+
        theme(aspect.ratio=1)

p_cluster + p_time + plot_annotation("Endothelium")

```


## Identify marker genes in each cluster
```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/endo_markers_clusters.rds"
if (!file.exists(filename)) {
  endo_markers <- parallelFindAllMarkers(endo)
  saveRDS(endo_markers, filename)
} else {
  endo_markers <- readRDS(filename)
}
```

### Cluster 0
```{r}
endo_markers[[0 + 1]]
```

### Cluster 1
```{r}
endo_markers[[1 + 1]]
```

### Cluster 2
```{r}
endo_markers[[2 + 1]]
```

### Cluster 3
```{r}
endo_markers[[3 + 1]]
```

### Cluster 4
```{r}
endo_markers[[4 + 1]]
```

### Cluster 5
```{r}
endo_markers[[5 + 1]]
```


## Use known marker genes to determine the identity of the clusters

Most of the markers are from here: https://elifesciences.org/articles/53072
```{r}


marker_genes <- c("Gpihbp1", "Kit", # Microvascular
                  "Car4", "Kdr", # Car4
                  "Mki67", "Top2a", # Proliferating miEC
                  "Vwf", "Vcam1", #macrovascular
                  "Vegfc", "Prss23", #Venous macrovascular
                  "Cxcl12", "Pcsk5", #Artearial macrovascular
                  "Ephb4", # Vein
                  "Flt4", "Ccl21a" # Lymphatic
)


heatmap_df <- make_heatmap_df(endo, marker_genes, sort_clusters = FALSE)

heatmap_df <- heatmap_df %>% group_by(gene,cluster) %>% summarise(expression_mean=mean(expression))

# For brackets and text (see ./helper_functions/brackets.R)
n_clusters <- max(as.numeric(Idents(endo)))
text_vec <- list(c("Microvascular", 2),
                 c("Car4+", 2),
                 c("Prolif.", 2),
                  c("maEC", 2),
                 c("Venous maEC", 2),
                 c("Arterial maEC", 2),
                 c("Venous", 1),
                 c("Lymphatic", 2)


)

endo_heatmap <- ggplot(heatmap_df, aes(x = gene, y = cluster, fill = expression_mean)) +
        geom_tile(color = "white", size = 0.1) +
        scale_fill_distiller(palette = "Blues", direction = 1, trans = "sqrt", name = "Expression") +
        coord_fixed(ratio = 1, xlim = NULL, ylim = c(1,n_clusters), expand = TRUE, clip = "off") +
        theme(plot.margin=unit(c(1.5,1,1,1),"cm")) +
        addBrackets(text_vec) +
        addText(text_vec, n_clusters) +
        theme(panel.background = element_rect(fill = "white", colour = "black", size = 0),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.title.x  = element_blank(),
              axis.text.x = element_text(size = 12, face = "italic"),
              axis.text.y = element_text(size = 12),
              axis.title.y = element_text(size = 14),
              legend.title = element_text(size = 14),
              legend.text = element_text(size = 12)
              ) +
        labs(y = "Cluster") +
        theme(axis.text.x = element_text(angle = 45, hjust=1))

endo_heatmap

pdf("./figures/endo/supp_cluster_markers.pdf", height = 5, width = 12)
endo_heatmap
dev.off()
```


## Fraction of the cells over time
```{r}
timepoint_factor <- factor(endo$timepoint)
timepoint_ordered <- ordered(timepoint_factor, levels = unique(endo$timepoint))

cell_ident_df <- data.frame(timepoint = timepoint_ordered,
                            cluster = Idents(endo))

prop_celltypes <- round(prop.table(table(cell_ident_df$timepoint, cell_ident_df$cluster), 1) * 100, 1)


ggplot(melt(prop_celltypes), aes(y=value, x=Var1)) +
        geom_area(aes(color = as.factor(Var2), group=as.factor(Var2), fill = as.factor(Var2))) +
        xlab("Day") +
        ylab("% of cells") +
        labs(fill = "Cluster", color = "Cluster", group = "Cluster") +
        scale_colour_manual(aesthetics = c("color", "fill"), values=color_category_20)
```


## Discard non-lung cell types, and combine similar cell types

Cells that are similar tend to cluster based on time, however we will group them together because they are truly the same cell type.

```{r}
endo_relabel <- endo
# Label the cell-types
levels(Idents(endo_relabel)) <- c("miEC", #0
                                  "Car4+", #1
                                  "Proliferating miEC", #2
                                  "Venous maEC", #3
                                  "Arterial maEC", #4
                                  "Lymphatic" #5

)
Idents(endo_relabel) <- ordered(Idents(endo_relabel), c("Arterial maEC", "Venous maEC", "miEC", "Car4+", "Proliferating miEC", "Lymphatic"))
endo_relabel$bulk_celltype <- "Endothelium"

filename <- "./data/endo_full_celltype_labels.rds"
if (!file.exists(filename)) {
  saveRDS(endo_relabel, filename)
} else {
  endo_relabel <- readRDS(filename)
}
```



# Relabel clusters based on marker gene expression
```{r, fig.width = 12}

renameFactorIdent <- function(obj, start, end){
  levels(obj)[levels(obj) == start] <- end
  return(obj)
}

Idents(endo_relabel) <- renameFactorIdent(Idents(endo_relabel), "Car4+", "*Car4*+")


p_rename_cluster <- DimPlot(endo_relabel) + umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_markdown(size=14),
              legend.title=element_text(size=14),
              plot.title = element_text(size = 14)
              )

p_time <- DimPlot(endo_relabel, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14),
              plot.title = element_text(size = 14))

p_time + p_rename_cluster + plot_annotation("Endothelium", theme = theme(plot.title = element_text(size = 14)))


pdf("./figures/endo/umap.pdf", width = 6, height = 4.5)
p_rename_cluster + plot_annotation("Endothelium")
dev.off()

pdf("./figures/endo/supp_umap.pdf", width = 18, height = 4.5)
p_cluster + p_time + p_rename_cluster + plot_annotation("Endothelium", theme = theme(plot.title = element_text(size = 14)))
dev.off()
```

## Re-confirm the markers in the relabeled clusters

```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/endo_markers_clusters_relabeled.rds"
if (!file.exists(filename)) {
  endo_relabeled_markers <- parallelFindAllMarkers(endo_relabel)
  saveRDS(endo_relabeled_markers, filename)
} else {
  endo_relabeled_markers <- readRDS(filename)
}
```


### Arterial maEC (macrovascular)
```{r}
endo_relabeled_markers[[0 + 1]]
```

### Venous maEC (macrovascular)
```{r}
endo_relabeled_markers[[0 + 2]]
```
### miEC (microvascular)
```{r}
endo_relabeled_markers[[0 + 3]]
```

### Car4+
```{r}
endo_relabeled_markers[[0 + 4]]
```

### Lymphatic
```{r}
endo_relabeled_markers[[0 + 5]]
```

### Proliferating miEC
```{r}
endo_relabeled_markers[[0 + 6]]
```


```{r}
timepoint_factor <- factor(endo_relabel$timepoint)
timepoint_ordered <- ordered(timepoint_factor, levels = unique(endo_relabel$timepoint))

cell_ident_df <- data.frame(timepoint = timepoint_ordered,
                            cluster = Idents(endo_relabel))

prop_celltypes <- round(prop.table(table(cell_ident_df$timepoint, cell_ident_df$cluster), 1) * 100, 1)


endo_mountain <- ggplot(melt(prop_celltypes), aes(y=value, x=Var1)) +
        geom_area(aes(color = as.factor(Var2), group=as.factor(Var2), fill = as.factor(Var2))) +
        xlab("Day") +
        ylab("Percent of cells") +
        labs(fill = "Cluster", color = "Cluster", group = "Cluster") +
        scale_colour_manual(aesthetics = c("color", "fill"), values=color_category_20) +
        theme(legend.key=element_blank(),
              legend.text=element_markdown(size=14),
              legend.title=element_text(size=14),
              axis.text.x  = element_text(size=14),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=14),
              axis.title.y = element_text(size=14),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_rect(fill = "white")
        ) +
        scale_x_discrete(expand = c(.01, .01)) +
        scale_y_continuous(expand = c(.01,.01)) +
        coord_fixed(0.05)

endo_mountain

pdf("./figures/endo/labeled_mountain.pdf", width = 7, height = 4)
endo_mountain
dev.off()

```

## Car4 expression across the endothelium
```{r, fig.width = 6}
car4_endo_plot <- FeaturePlot(endo, "Car4") +
        featureplot_theme() +
        theme(aspect.ratio=1,
              plot.title = element_text(face = "italic", size = 14),
              legend.title = element_text(size = 14),
              legend.text = element_text(size = 12)
        ) +
        scale_color_viridis(name = "Expression", direction = -1)

car4_endo_plot

pdf("./figures/endo/endo_car4_plot.pdf", width = 6, height = 4.5)
car4_endo_plot
dev.off()
```

## Kdr expression across the endothelium
```{r, fig.width = 6}
kdr_endo_plot <- FeaturePlot(endo, "Kdr", pt.size = 0.5) +
        featureplot_theme() +
        theme(aspect.ratio=1,
              plot.title = element_text(face = "italic", size = 18),
              legend.title = element_text(size = 14),
              legend.text = element_text(size = 12)
        ) +
        scale_color_viridis(name = "Expression", direction = -1)

kdr_endo_plot

pdf("./figures/endo/endo_car4_plot.pdf", width = 6, height = 4.5)
kdr_endo_plot
dev.off()
```

## Integrating velocity measures


```{r}
endo_relabel$car4_prob <- readRDS("./velocity/endo/data/endo_car4_prob.rds")$to_Car4
```
```{r}
FeaturePlot(subset(subset(endo_relabel, car4_prob > 0.2), idents = "Car4+", invert = TRUE), "car4_prob") +
        featureplot_theme() +
        theme(aspect.ratio=1) +
        scale_color_viridis(name = "Expression", direction = 1)
```

```{r}
endo_relabel$car4_originators <- "no"
endo_relabel$car4_originators[endo_relabel$car4_prob > 0.2 & Idents(endo_relabel) != "Car4"] <- "yes"
```

```{r}
car4_originator_markers <- FindMarkers(endo_relabel, group.by = "car4_originators", "yes", "no", test.use = "MAST")
```

```{r}
car4_originator_markers %>% arrange(-avg_logFC)
write.xlsx(car4_originator_markers, file = "./figures/endo/Table S2 - Commitment to Car4.xlsx", rowNames = TRUE)
```