---
title: "scSEQ analysis of the developing mesenchyme"
author: Nick Negretti
date: 11/17/20
output: rmarkdown::github_document
---

# Analysis of the lung mesenchyme

## Load libraries and helper functions
```{r, results="hide", message = FALSE}
setwd("~/postdoc/code/devo_scseq_github")
source("./helper_functions/globals.R")
source("./helper_functions/libraries.R")

opts_knit$set(root.dir = getwd())

source("./helper_functions/trajectory.R")
source("./helper_functions/cluster.R")
source("./helper_functions/colors.R")
source("./helper_functions/brackets.R")
source("./helper_functions/heatmaps.R")

plan("multiprocess", workers = N_WORKERS)
```
## Load data from merge and clean pipeline
```{r}
meso <- readRDS("./data/meso_full_noGm42418_sct_p7b_integrated_retransform.rds")
```

## UMAP and clustering
```{r, fig.width = 12}
plan("sequential")

# Make the tiempoints an ordered factor, so they are displayed consistantly
meso$timepoint <- ordered(as.factor(meso$timepoint), unique(meso$timepoint))

# cluster and UMAP from the previous processing seems fine
# k_param = 25, dims_neighbors = 1:15, cluster_res = 0.15  - is what is after the integration pipeline

p_cluster <- DimPlot(meso) + umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14))
p_time <- DimPlot(meso, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_scanpy_default)+
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14))

p_cluster + p_time + plot_annotation("Mesenchyme")


saveTiff("./figures/meso/supp_umap.tiff",
         p_cluster + p_time + plot_annotation("Mesenchyme"),
         width = 12, height = 4.5)

```

## Identify marker genes in each cluster
```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/meso_markers_clusters.rds"
if (!file.exists(filename)) {
  meso_markers <- parallelFindAllMarkers(meso)
  saveRDS(meso_markers, filename)
} else {
  meso_markers <- readRDS(filename)
}
```

### Cluster 0
```{r}
meso_markers[[0 + 1]][n_print,]
```

### Cluster 1
```{r}
meso_markers[[1 + 1]][n_print,]
```

### Cluster 2
```{r}
meso_markers[[2 + 1]][n_print,]
```

### Cluster 3
```{r}
meso_markers[[3 + 1]][n_print,]
```

### Cluster 4
```{r}
meso_markers[[4 + 1]][n_print,]
```

### Cluster 5
```{r}
meso_markers[[5 + 1]][n_print,]
```

### Cluster 6
```{r}
meso_markers[[6 + 1]][n_print,]
```

### Cluster 7
```{r}
meso_markers[[7 + 1]][n_print,]
```

### Cluster 8
```{r}
meso_markers[[8 + 1]][n_print,]
```


## Use known marker genes to determine the identity of the clusters

```{r}


marker_genes <- c("Wnt2", "Macf1", # Wnt2 fibroblasts 0
                  "Tgfbi", "Wnt5a", # Myofibroblasts 1
                  "Dcn", "Col1a1", # Adventitial fibroblast 2
                  "Top2a", "Mki67", # Prenatal Wnt2 3
                  "Cspg4", "Cox4i2", # Pericyte 4
                  "Ednrb", "Pdgfra", # Prenatal myo 5
                  "Wt1", "Upk3b", # Mesothelium 6
                  "Eln", "Acta2", # Smooth muscle 7
                  "Mpz", "Mal", # Neuron 8
                  "Tnnt2", "Actc1" # Cardiomyocyte
)


heatmap_df <- make_heatmap_df(meso, marker_genes, sort_clusters = FALSE)

heatmap_df <- heatmap_df %>% group_by(gene,cluster) %>% summarise(expression_mean=mean(expression))

# For brackets and text (see ./helper_functions/brackets.R)
n_clusters <- max(as.numeric(Idents(meso)))
text_vec <- list(c("Wnt2+ FB", 2),
                 c("Myofibroblast", 2),
                 c("Adventitial FB", 2),
                 c("Prenatal\nWnt2+ FB", 2),
                 c("Pericyte", 2),
                 c("Prenatal\nMyofibroblast", 2),
                 c("Mesothelium", 2),
                 c("Smooth\nmuscle", 2),
                 c("Neuron", 2),
                 c("Cardiomyocyte", 2)
)

meso_heatmap <- ggplot(heatmap_df, aes(x = gene, y = cluster, fill = expression_mean)) +
        geom_tile(color = "white", size = 0.1) +
        scale_fill_distiller(palette = "Blues", direction = 1, trans = "sqrt", name = "Expression") +
        coord_fixed(ratio = 1, xlim = NULL, ylim = c(1,n_clusters), expand = TRUE, clip = "off") +
        theme(plot.margin=unit(c(1.5,1,1,1),"cm")) +
        addBrackets(text_vec) +
        addText(text_vec, n_clusters) +
        theme(panel.background = element_rect(fill = "white", colour = "black", size = 0),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.title.x  = element_blank(),
              axis.text.x = element_text(size = 12, face = "italic"),
              axis.text.y = element_text(size = 12),
              axis.title.y = element_text(size = 14),
              legend.title = element_text(size = 14),
              legend.text = element_text(size = 12)
        ) +
        labs(y = "Cluster") +
        theme(axis.text.x = element_text(angle = 45, hjust=1))

meso_heatmap


saveTiff("./figures/meso/supp_cluster_markers.tiff",
         meso_heatmap,
         width = 12, height = 5.5)

```


## Fraction of the cells over time
```{r}
timepoint_factor <- factor(meso$timepoint)
timepoint_ordered <- ordered(timepoint_factor, levels = unique(meso$timepoint))

cell_ident_df <- data.frame(timepoint = timepoint_ordered,
                            cluster = Idents(meso))

prop_celltypes <- round(prop.table(table(cell_ident_df$timepoint, cell_ident_df$cluster), 1) * 100, 1)


ggplot(melt(prop_celltypes), aes(y=value, x=Var1)) +
        geom_area(aes(color = as.factor(Var2), group=as.factor(Var2), fill = as.factor(Var2))) +
        xlab("Day") +
        ylab("% of cells") +
        labs(fill = "Cluster", color = "Cluster", group = "Cluster") +
        scale_colour_manual(aesthetics = c("color", "fill"), values=color_category_20)
```



## Apply labels

```{r}
meso_relabel <- meso
# Label the cell-types
text_vec <- list(c("Wnt2+ FB", 2),
                 c("Myofibroblast", 2),
                 c("Adventitial", 2),
                 c("Prolif.\nWnt2+ FB", 2),
                 c("Pericyte", 2),
                 c("Prolif.\nMyofibroblast", 2),
                 c("Mesothelium", 2),
                 c("Smooth\nmuscle", 2),
                 c("Neuron", 2))
levels(Idents(meso_relabel)) <- c("Wnt2+ FB", #0
                                  "Myofibroblast", #1
                                  "Adventitial fibroblast", #2
                                  "Proliferating Wnt2+ FB", #3
                                  "Pericyte", #4
                                  "Proliferating myofibroblast", #5
                                  "Mesothelium", #6
                                  "Smooth muscle", #7
                                  "Neuron" #8

)
Idents(meso_relabel) <- ordered(Idents(meso_relabel), c("Proliferating Wnt2+ FB", "Wnt2+ FB", "Proliferating myofibroblast", "Myofibroblast", "Adventitial fibroblast", "Pericyte", "Mesothelium", "Smooth muscle", "Neuron"))

Ident_md <- Idents(meso_relabel)
levels(Ident_md)[1:2] <- c("Prenatal *Wnt2*+ FB", "*Wnt2*+ FB")

meso_relabel$Ident_md <- Ident_md

meso_relabel$bulk_celltype <- "Mesenchyme"

# Save this for use by other stages of the analysis
filename <- "./data/meso_full_celltype_labels.rds"
if (!file.exists(filename)) {
  saveRDS(meso_relabel, filename)
}
```


# Relabel clusters based on marker gene expression
```{r, fig.width = 12}

p_rename_cluster <- DimPlot(meso_relabel, group.by = "Ident_md") + umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_markdown(size=14),
              legend.title=element_text(size=14),
              plot.title = element_text(size=14)) +
        ggtitle("Mesenchyme")


p_time <- DimPlot(meso_relabel, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14))

p_time + p_rename_cluster + plot_annotation("Mesenchyme")



saveTiff("./figures/meso/umap.tiff",
         p_rename_cluster,
         width = 6, height = 4.5)

```

## Re-confirm the markers in the relabeled clusters

```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/meso_markers_clusters_relabeled.rds"
if (!file.exists(filename)) {
  meso_relabeled_markers <- parallelFindAllMarkers(meso_relabel)
  saveRDS(meso_relabeled_markers, filename)
} else {
  meso_relabeled_markers <- readRDS(filename)
}
```


### Prenatal Wnt2+
```{r}
meso_relabeled_markers[[0 + 1]][n_print,]
```

### Wnt2+
```{r}
meso_relabeled_markers[[1 + 1]][n_print,]
```
### Prenatal Myofibroblast
```{r}
meso_relabeled_markers[[2 + 1]][n_print,]
```

### Myofibroblast
```{r}
meso_relabeled_markers[[3 + 1]][n_print,]
```

### Adventitial fibroblast
```{r}
meso_relabeled_markers[[4 + 1]][n_print,]
```

### Pericyte
```{r}
meso_relabeled_markers[[5 + 1]][n_print,]
```

### Mesothelium
```{r}
meso_relabeled_markers[[6 + 1]][n_print,]
```

### Neuron
```{r}
meso_relabeled_markers[[7 + 1]][n_print,]
```


```{r}
timepoint_factor <- factor(meso_relabel$timepoint)
timepoint_ordered <- ordered(timepoint_factor, levels = unique(meso_relabel$timepoint))

cell_ident_df <- data.frame(timepoint = timepoint_ordered,
                            cluster = Idents(meso_relabel))

prop_celltypes <- round(prop.table(table(cell_ident_df$timepoint, cell_ident_df$cluster), 1) * 100, 1)

print(prop_celltypes)
colnames(prop_celltypes)[1:2] <- c("Prenatal *Wnt2*+ FB", "*Wnt2*+ FB")
meso_mountain <- ggplot(melt(prop_celltypes), aes(y=value, x=Var1)) +
        geom_area(aes(color = as.factor(Var2), group=as.factor(Var2), fill = as.factor(Var2))) +
        xlab("Day") +
        ylab("% of cells") +
        labs(fill = "Cluster", color = "Cluster", group = "Cluster") +
        scale_colour_manual(aesthetics = c("color", "fill"), values=color_category_20) +
        theme(legend.key = element_blank(),
              legend.text = element_markdown(size=14),
              legend.title = element_text(size=14),
              axis.text.x  = element_text(size=14),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=14),
              axis.title.y = element_text(size=14),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_rect(fill = "white")
        ) +
        scale_x_discrete(expand = c(.01, .01)) +
        scale_y_continuous(expand = c(.01,.01)) +
        coord_fixed(0.05)


meso_mountain

saveTiff("./figures/meso/labeled_mountain.tiff",
         meso_mountain,
         width = 9, height = 4)

```

## Wnt2 and Wnt5a expression
```{r, fig.width = 12}
wnt2_meso_plot <- FeaturePlot(meso, "Wnt2") +
        featureplot_theme() +
        theme(aspect.ratio=1) +
        scale_color_viridis(name = "Expression", direction = -1)
wnt5a_meso_plot <- FeaturePlot(meso, "Wnt5a") +
        featureplot_theme() +
        theme(aspect.ratio=1) +
        scale_color_viridis(name = "Expression", direction = -1)

wnt2_meso_plot + wnt5a_meso_plot + plot_layout(ncol = 2)

saveTiff("./figures/meso/meso_wnt_plot.tiff",
         wnt2_meso_plot + wnt5a_meso_plot + plot_layout(ncol = 2),
         width = 12, height = 4.5)

```


# Higher res clustering

## UMAP and clustering with increased granularity
```{r, fig.width = 12}
plan("sequential")

# cluster and UMAP from the previous processing seems fine
# k_param = 25, dims_neighbors = 1:15, cluster_res = 0.15  - is what is after the integration pipeline
meso_highres <- cluster_pca_umap(meso, k_param = 25, dims_umap = 1:15, dims_neighbors = 1:25, cluster_res = 0.2) # Note, this breaks if future is set to plan: multiprocess


p_cluster <- DimPlot(meso_highres) + umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_category_20) +
        theme(aspect.ratio=1)
p_time <- DimPlot(meso_highres, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_scanpy_default)+
        theme(aspect.ratio=1)

p_cluster + p_time + plot_annotation("Mesenchyme")

#pdf("./figures/meso/supp_umap_highres.pdf", width = 12, height = 4.5)
#p_cluster + p_time + plot_annotation("Mesenchyme")
#dev.off()
```

```{r}

ggplot_data <- as.data.frame(Embeddings(meso_highres, reduction = "umap"))
ggplot_data$named_clusters <- Idents(meso_relabel)
ggplot_data$timepoint <- meso_relabel$timepoint
ggplot_data$subclusters <- Idents(meso_highres)

myo_highlight <- ggplot() +
        geom_point(data = ggplot_data[!(ggplot_data$named_clusters %in% c("Proliferating myofibroblast", "Myofibroblast")),],
                   aes(x = UMAP_1, y = UMAP_2), size = 0.1, alpha = 0.2, color = "gray") +
        geom_point(data = ggplot_data[(ggplot_data$named_clusters %in% c("Proliferating myofibroblast", "Myofibroblast")),],
                   aes(x = UMAP_1, y = UMAP_2, color = timepoint), size = 0.1) +
        umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_scanpy_default) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14)) +
        guides(colour = guide_legend(override.aes = list(size=3))) +
        ggtitle("Myofibroblasts by time")

myo_highlight


myo_supp_highlight <- ggplot() +
        geom_point(data = ggplot_data[!(ggplot_data$subclusters %in% c(0,6,8,10)),],
                   aes(x = UMAP_1, y = UMAP_2), size = 0.1, alpha = 0.2, color = "gray") +
        geom_point(data = ggplot_data[(ggplot_data$subclusters %in% c(0,6,8,10)),],
                   aes(x = UMAP_1, y = UMAP_2, color = subclusters), size = 0.1) +
        umap_theme() +
        scale_colour_manual(name = "Cluster", values = color_scanpy_default, labels = c("E12/E15 Myofibroblast",
                                                                                        "E18 Myofibroblast",
                                                                                        "Replicating myofibroblast",
                                                                                        "Mature myofibroblast"),
        breaks = c(10,6,8,0)) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14)) +
        guides(colour = guide_legend(override.aes = list(size=3))) +
        ggtitle("Myofibroblast subclusters")

myo_supp_highlight

saveTiff("./figures/meso/umap_myo_timepoint.tiff",
         myo_highlight,
         width = 5, height = 4.5)


saveTiff("./figures/meso/umap_myo_supp_cluster.tiff",
         myo_supp_highlight,
         width = 6, height = 4.5)


```

## Identify marker genes in each cluster
```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/meso_markers_clusters_highres.rds"
if (!file.exists(filename)) {
  meso_markers_highres <- parallelFindAllMarkers(meso_highres)
  saveRDS(meso_markers_highres, filename)
} else {
  meso_markers_highres <- readRDS(filename)
}
```

### Cluster 0
```{r}
meso_markers_highres[[0 + 1]][n_print,]
```

### Cluster 1
```{r}
meso_markers_highres[[1 + 1]][n_print,]
```

### Cluster 2
```{r}
meso_markers_highres[[2 + 1]][n_print,]
```

### Cluster 3
```{r}
meso_markers_highres[[3 + 1]][n_print,]
```

### Cluster 4
```{r}
meso_markers_highres[[4 + 1]][n_print,]
```

### Cluster 5
```{r}
meso_markers_highres[[5 + 1]][n_print,]
```

### Cluster 6
```{r}
meso_markers_highres[[6 + 1]][n_print,]
```

### Cluster 7
```{r}
meso_markers_highres[[7 + 1]][n_print,]
```

### Cluster 8
```{r}
meso_markers_highres[[8 + 1]][n_print,]
```

### Cluster 9
```{r}
meso_markers_highres[[9 + 1]][n_print,]
```

### Cluster 10
```{r}
meso_markers_highres[[10 + 1]][n_print,]
```

### Cluster 11
```{r}
meso_markers_highres[[11 + 1]][n_print,]
```

### Cluster 12
```{r}
meso_markers_highres[[12 + 1]][n_print,]
```

# Save markers for detailed meso clusters
```{r}
meso_highres_myo_clusters <- c(0,6,8,10)
meso_highres_myo_names <- c("Mature myofibroblasts", "E18 myofibroblasts", "Proliferating myofibroblasts", "E12-E15 myofibroblasts")

wb_myo_markers <- createWorkbook()
for (idx in seq_along(meso_highres_myo_clusters)){
  i <- meso_highres_myo_clusters[idx] + 1
  addWorksheet(wb_myo_markers, meso_highres_myo_names[idx])
  writeData(wb_myo_markers, meso_highres_myo_names[idx], meso_markers_highres[[i]], rowNames = TRUE)
}
saveWorkbook(wb_myo_markers, file = "./figures/meso/myo_highres_marker_genes.xlsx", overwrite = TRUE)


```

```{r}
wnt5a_all_myo <- VlnPlot(subset(meso_relabel, ident = c("Myofibroblast", "Prenatal Myofibroblast")), "Wnt5a", group.by = "timepoint") +ggtitle("All myofibroblasts")
wnt5a_mature_myo <-  VlnPlot(subset(meso_highres, ident = 0), "Wnt5a", group.by = "timepoint") + ggtitle("Mature myofibroblasts")
wnt5a_proliferating_myo <- VlnPlot(subset(meso_highres, ident = 8), "Wnt5a", group.by = "timepoint") + ggtitle("Proliferating myofibroblasts")
wnt5a_e18_myo <- VlnPlot(subset(meso_highres, ident = 6), "Wnt5a", group.by = "timepoint") + ggtitle("'E18' myofibroblasts")
```