---
title: "scSEQ analysis of the developing epithelium"
author: Nick Negretti
date: 11/17/20
output: rmarkdown::github_document
---

# Analysis of the lung epithelium

## Load libraries and helper functions
```{r, results="hide", message = FALSE}
setwd("~/postdoc/code/devo_scseq")

source("./helper_functions/libraries.R")

opts_knit$set(root.dir = getwd())

source("./helper_functions/trajectory.R")
source("./helper_functions/cluster.R")
source("./helper_functions/colors.R")
source("./helper_functions/brackets.R")
source("./helper_functions/heatmaps.R")
set.seed(42) # For reproducability

N_WORKERS <- 3
options(future.globals.maxSize=2048*1024^2)
plan("multiprocess", workers = N_WORKERS)
```
## Load data from merge and clean pipeline
```{r}
epi <- readRDS("./data/epi_full_noGm42418_sct_p7b_integrated_retransform.rds")
epi
```


## UMAP and clustering
```{r, fig.width = 12}
plan("sequential")

# Make the tiempoints an ordered factor, so they are displayed consistantly
epi$timepoint <- ordered(as.factor(epi$timepoint), unique(epi$timepoint))

epi <- cluster_pca_umap(epi, k_param = 5, dims_umap = c(1:5,7:19), dims_neighbors = c(1:5,7:30), cluster_res = 0.2) # Note, this breaks if future is set to plan: multiprocess

levels(Idents(epi)) <- c(0, 1, 0, 0, 2, 0, 3, 0, 4, 0, 5, 6)

p_cluster <- DimPlot(epi) + umap_theme() +
        scale_colour_manual(name = "Louvain Cluster", values = color_category_20) +
        theme(aspect.ratio=1)
p_time <- DimPlot(epi, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_scanpy_default)+
        theme(aspect.ratio=1)

p_cluster + p_time + plot_annotation("Epithelium")

pdf("./figures/epi/supp_umap.pdf", width = 12, height = 4.5)
p_cluster + p_time + plot_annotation("Epithelium")
dev.off()
```

```{r}
png("./epithelium/PCs.png", width = 800, height = 4048, res = 150)
DefaultAssay(epi) <- "integrated"
DimHeatmap(epi, dims = 1:30, nfeatures = 22)
DefaultAssay(epi) <- "SCT"
dev.off()
```

## Identify marker genes in each cluster
```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/epi_markers_clusters.rds"
if (!file.exists(filename)) {
  epi_markers <- parallelFindAllMarkers(epi)
  saveRDS(epi_markers, filename)
} else {
  epi_markers <- readRDS(filename)
}
```

### Cluster 0
```{r}
epi_markers[[0 + 1]]
```

### Cluster 1
```{r}
epi_markers[[1 + 1]]
```

### Cluster 2
```{r}
epi_markers[[2 + 1]]
```

### Cluster 3
```{r}
epi_markers[[3 + 1]]
```

### Cluster 4
```{r}
epi_markers[[4 + 1]]
```

### Cluster 5
```{r}
epi_markers[[5 + 1]]
```

### Cluster 6
```{r}
epi_markers[[6 + 1]]
```

## Use known marker genes to determine the identity of the clusters
```{r}


marker_genes <- c("Sftpa1", "Sftpc", # AT2
                  "Hopx", "Aqp5", #AT1
                  "Foxj1", "Dynlrb2", # Cilliated
                  "Mdk", "Mki67", # Primordial
                  "Scgb1a1", "Scgb3a2", # Secretory
                  "Cdkn1a", "Cldn4", # Transitional?
                  "Ascl1", "Scg5" #Neuroendocrine
                 )


heatmap_df <- make_heatmap_df(epi, marker_genes, sort_clusters = FALSE)

heatmap_df <- heatmap_df %>% group_by(gene,cluster) %>% summarise(expression_mean=mean(expression))

# For brackets and text (see ./helper_functions/brackets.R)
n_clusters <- max(as.numeric(Idents(epi)))
text_vec <- list(c("AT2", 2),
                 c("AT1", 2),
                 c("Cilliated", 2),
                 c("Primordial", 2),
                 c("Secretory", 2),
                 c("Transitional", 2),
                 c("Neuroendocrine", 2)
)

epi_heatmap <- ggplot(heatmap_df, aes(x = gene, y = cluster, fill = expression_mean)) +
  geom_tile(color = "white", size = 0.1) +
  scale_fill_distiller(palette = "Blues", direction = 1, trans = "sqrt", name = "Expression") +
  coord_fixed(ratio = 1, xlim = NULL, ylim = c(1,n_clusters), expand = TRUE, clip = "off") +
  theme(plot.margin=unit(c(1.5,1,1,1),"cm")) +
        addBrackets(text_vec) +
        addText(text_vec, n_clusters) +
  theme(panel.background = element_rect(fill = "white", colour = "black", size = 0),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.x  = element_blank()) +
  labs(y = "Cluster") +
  theme(axis.text.x = element_text(angle = 45, hjust=1, face = "italic"))

epi_heatmap

pdf("./figures/epi/supp_cluster_markers.pdf", height = 5, width = 12)
epi_heatmap
dev.off()

```

## Fraction of the cells over time
```{r}
timepoint_factor <- factor(epi$timepoint)
timepoint_ordered <- ordered(timepoint_factor, levels = unique(epi$timepoint))

cell_ident_df <- data.frame(timepoint = timepoint_ordered,
                            cluster = Idents(epi))

prop_celltypes <- round(prop.table(table(cell_ident_df$timepoint, cell_ident_df$cluster), 1) * 100, 1)


ggplot(melt(prop_celltypes), aes(y=value, x=Var1)) +
        geom_area(aes(color = as.factor(Var2), group=as.factor(Var2), fill = as.factor(Var2))) +
        xlab("Day") +
        ylab("% of cells") +
        labs(fill = "Cluster", color = "Cluster", group = "Cluster") +
        scale_colour_manual(aesthetics = c("color", "fill"), values=color_category_20)
```


## Name cell types

Cells that are similar tend to cluster based on time, however we will group them together because they are truly the same cell type.

```{r}
epi_relabel <- epi
# Label the cell-types
levels(Idents(epi_relabel)) <- c("AT2", #0
                                  "AT1", #1
                                  "Cilliated", #2
                                  "Primordial", #3
                                  "Secretory", #4
                                  "Transitional", #5
                                  "Neuroendocrine" #6

)
Idents(epi_relabel) <- ordered(Idents(epi_relabel), c("Primordial", "Transitional", "AT1", "AT2", "Cilliated", "Secretory", "Neuroendocrine"))
epi_relabel$bulk_celltype <- "Epithelium"
#saveRDS(epi_relabel, "./data/epi_full_celltype_labels.rds")
```



# Relabel clusters based on marker gene expression
```{r, fig.width = 12}
p_rename_cluster <- DimPlot(epi_relabel) + umap_theme() +
  scale_colour_manual(name = "Cluster", values = color_category_20) +
  theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14))

p_time <- DimPlot(epi_relabel, group.by = "timepoint") + umap_theme() +
        scale_colour_manual(name = "Timepoint", values = color_category_20) +
        theme(aspect.ratio=1) +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14))

p_time + p_rename_cluster + plot_annotation("Epithelium")

pdf("./figures/epi/umap.pdf", width = 6, height = 4.5)
p_rename_cluster + plot_annotation("Epithelium")
dev.off()

pdf("./figures/epi/supp_umap.pdf", width = 18, height = 4.5)
p_cluster + p_time + p_rename_cluster + plot_annotation("Epithelium")
dev.off()
```

## Re-confirm the markers in the relabeled clusters

```{r}
plan("multiprocess", workers = N_WORKERS)
filename <- "./data/epi_markers_clusters_relabeled.rds"
if (!file.exists(filename)) {
  epi_relabeled_markers <- parallelFindAllMarkers(epi_relabel)
  saveRDS(epi_relabeled_markers, filename)
} else {
  epi_relabeled_markers <- readRDS(filename)
}
```



### Primordial
```{r}
epi_relabeled_markers[[0 + 1]]
```

### AT1
```{r}
epi_relabeled_markers[[1 + 1]]
```

### AT2
```{r}
epi_relabeled_markers[[2 + 1]]
```
### Cilliated
```{r}
epi_relabeled_markers[[3 + 1]]
```

### Secretory
```{r}
epi_relabeled_markers[[4 + 1]]
```

### Transitional
```{r}
epi_relabeled_markers[[5 + 1]]
```
### Neuroendocrine
```{r}
epi_relabeled_markers[[6 + 1]]
```

```{r}
timepoint_factor <- factor(epi_relabel$timepoint)
timepoint_ordered <- ordered(timepoint_factor, levels = unique(epi_relabel$timepoint))

cell_ident_df <- data.frame(timepoint = timepoint_ordered,
                            cluster = Idents(epi_relabel))

prop_celltypes <- round(prop.table(table(cell_ident_df$timepoint, cell_ident_df$cluster), 1) * 100, 1)


epi_mountain <- ggplot(melt(prop_celltypes) %>% arrange(Var2), aes(y=value, x=Var1)) +
        geom_area(aes(color = as.factor(Var2), group=as.factor(Var2), fill = as.factor(Var2))) +
        xlab("Day") +
        ylab("% of cells") +
        labs(fill = "Cluster", color = "Cluster", group = "Cluster") +
        scale_colour_manual(aesthetics = c("color", "fill"), values=color_category_20) +
        theme(legend.key=element_blank(),
              legend.text=element_text(size=14),
              legend.title=element_text(size=14),
              axis.text.x  = element_text(size=14),
              axis.text.y = element_text(size=14),
              axis.title.x = element_text(size=14),
              axis.title.y = element_text(size=14),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.background = element_rect(fill = "white")
        ) +
        scale_x_discrete(expand = c(.01, .01)) +
        scale_y_continuous(expand = c(.01,.01)) +
        coord_fixed(0.05)

epi_mountain

pdf("./figures/epi/labeled_mountain.pdf", width = 7, height = 4)
epi_mountain
dev.off()
```

## Cdkn1a expression across the epithelium
```{r, fig.width = 6}
FeaturePlot(epi, "Cdkn1a") +
  featureplot_theme() +
  theme(aspect.ratio=1) +
  scale_color_viridis(name = "Expression", direction = -1)
```


## Diving into the expression patterns of the Cdkn1a cells (excluding the airway)
```{r}


cdkn1a_high_epi <- subset(subset(epi_relabel, Cdkn1a > 0.5), idents = "Cilliated", invert = TRUE)

marker_genes <- c("Hopx", "Ager", #AT1
                  "Sftpa1", "Sftpb", # AT2
                  "Mki67", "Cdkn1a" # Progenitors?
)

heatmap_df <- make_heatmap_df(cdkn1a_high_epi, marker_genes, sort_clusters = FALSE)

heatmap_df <- heatmap_df %>% group_by(gene,time) %>% summarise(expression_mean=mean(expression))


n_clusters <- max(as.numeric(Idents(cdkn1a_high_epi)))
#grid.brackets()
# Look in to grid.brackets to customize the brackets. Can prob make a function for this.
cdkn1a_heatmap <- ggplot(heatmap_df, aes(x = gene, y = time, fill = expression_mean)) +
        geom_tile(color = "white", size = 0.1) +
        scale_fill_distiller(palette = "Blues", direction = 1, trans = "sqrt", name = "Expression") +
        #ggtitle("Expression") +
        coord_fixed(ratio = 1) +
        theme(panel.background = element_rect(fill = "white"),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.title.x  = element_blank(),
              legend.text=element_text(size=14),
              legend.title=element_text(size=14),
              axis.text.x  = element_text(size=14, angle = 45, vjust = 1, hjust=1, face = "italic"),
              axis.text.y = element_text(size=14),
              axis.title.y = element_text(size=14),
              legend.key.size = unit(0.8, "cm"),
              plot.title = element_text(size=14)
        ) +
        labs(y = "Cluster") +
        ggtitle("Expression patterns in transitional cells \n(excluding airway)")

cdkn1a_heatmap

pdf("./figures/epi/cdkn1a_heatmap.pdf", height = 5, width = 5)
cdkn1a_heatmap
dev.off()
```

# Type 1 cells are not silent players

Drop E15 - it's unclear whether there are actually AT1 cells present that early.
```{r}


marker_genes <- c('Eln', 'Fbn1', 'Fbn2', 'Fbln1', 'Fbln2', 'Fbln3', 'Fbln4', 'Fbln5',
                  'Emilin1', 'Magp1', 'Magp2', 'Ltbp1', 'Ltbp2', 'Ltbp3', 'Ltbp4', 'Lox',
                  'Loxl1', 'Loxl2', 'Loxl3', 'Loxl4')


heatmap_df <- make_heatmap_df(subset(subset(epi_relabel, idents = "AT1"), timepoint %in% c("E12", "E15"), invert = TRUE), marker_genes, sort_clusters = FALSE)

heatmap_df <- heatmap_df %>% group_by(gene,time) %>% summarise(expression_mean=mean(expression))

addBackZeros <- function(df, markers){
  out <- do.call(rbind, lapply(markers[!(markers %in% unique(df$gene))], function(x){
    t(do.call(cbind, list(sapply(unique(levels(df$time)), function(y){
      c(x, y, 0)
    }))))
  }))
  colnames(out) <- colnames(df)
  out <- as.data.frame(out)
  out$expression_mean <- as.double(out$expression_mean)
  out$time <- ordered(out$time, levels(df$time))
  rbind(df, out)
}

n_clusters <- max(as.numeric(Idents(subset(epi_relabel, idents = "AT1"))))
#grid.brackets()
# Look in to grid.brackets to customize the brackets. Can prob make a function for this.
at1_matrix_plot <- ggplot(addBackZeros(heatmap_df, marker_genes), aes(x = gene, y = time, fill = expression_mean)) +
        geom_tile(color = "white", size = 0.1) +
        scale_fill_distiller(palette = "Blues", direction = 1, trans = "sqrt", name = "Expression") +
        #ggtitle("Expression") +
        coord_fixed(ratio = 0.6) +
        theme(panel.background = element_rect(fill = "white"),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.title.x  = element_blank(),
              legend.text=element_text(size=14),
              legend.title=element_text(size=14),
              axis.text.x  = element_text(size=14, angle = 45, vjust = 1, hjust=1, face = "italic"),
              axis.text.y = element_text(size=14),
              axis.title.y = element_text(size=14),
              legend.key.size = unit(0.8, "cm"),
              plot.title = element_text(size=14)
        ) +
        labs(y = "Timepoint") +
        ggtitle("Expression patterns in AT1 cells")


at1_matrix_plot

pdf("./figures/epi/epi_labeled_umap.pdf", width = 6, height = 4.5)
p_rename_cluster + plot_annotation("Epithelium")
dev.off()

pdf("./figures/epi/AT1_matrix.pdf", width = 12, height = 4)
at1_matrix_plot
dev.off()

```

```{r}

cdkn1a_epi_plot <- FeaturePlot(epi, "Cdkn1a", pt.size = 0.6) +
        featureplot_theme() +
        theme(aspect.ratio=1) +
        scale_color_viridis(name = "Expression", direction = -1, trans = "sqrt") +
        theme(legend.text=element_text(size=14),
              legend.title=element_text(size=14),
              legend.key.size = unit(0.8, "cm"),
        plot.title = element_text(face = "italic"))


cdkn1a_epi_plot

pdf("./figures/epi/epi_cdkn1a.pdf", width = 6, height = 4.5)
cdkn1a_epi_plot
dev.off()
```


## Transitional cluster expression patterns

```{r, fig.width = 12, fig.height = 15}


markers <- factor(c("Cdkn1a", "Cldn4", # Transitional?
                    "Hopx", "Aqp5", #AT1
                    "Sftpa1", "Sftpc", # AT2
                    "Foxj1", "Dynlrb2", # Cilliated
                    "Scgb1a1", "Scgb3a2", # Secretory
                    "Ascl1", "Scg5" # Neuroendocrine
                 ), ordered = TRUE
)


heatmap_df <- make_heatmap_df(epi_relabel, markers, sort_clusters = FALSE)

heatmap_df <- heatmap_df %>% group_by(cluster,gene) %>% summarise(expression_mean=mean(expression),
                                                                  frac_pos=(sum(expression > 0) * 100 / length(expression)))


heatmap_df <- heatmap_df %>% mutate(gene = factor(gene, markers))


transitional_heatmap <- ggplot(heatmap_df, aes(x = cluster, y = gene, color = expression_mean, size = frac_pos)) +
        geom_point(shape = 16) +
        scale_size(range = c(1, 9), name = "% expressing") +
        scale_color_distiller(palette = "Blues", direction = 1, trans = "sqrt", name = "Expression") +
        coord_fixed(ratio = 1) +
        theme(plot.margin=unit(c(1.5,1,1,1),"cm")) +
        theme(panel.background = element_rect(fill = "white", color = "black", size = 0.5),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              axis.title.x  = element_blank(),) +
        labs(y = "") +
        theme(axis.text.x  = element_text(size=14, angle = 45, vjust = 1, hjust=1),
              #axis.ticks.x = element_blank(),
              axis.text.y = element_text(size = 12, face = "italic"))


transitional_heatmap


pdf("./figures/epi/transitional_cluster_expression.pdf", width = 5.5, height = 5.5)
transitional_heatmap
dev.off()



```
